Normalization by Evaluation for Sized Dependent Types

### 2017-02-23 nbe/icfp17.tex
[ ICFP 17 ] Pi-injectivity, subject reduction, algorithmic subtyping.

\begin{lemma}[Inversion of typing]
\label{lem:invty} \bla
\begin{enumerate}
\item TODO
\end{enumerate}
\end{lemma}

@@

We can now define the normalization relation
$\nbe \Gamma T t \evalsto v \iff \R {|\Gamma|} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.

@@

\begin{lemma}[Identity environment]
  If $\der \Gamma$ then
  $\LRS \Gamma \sid \Gamma \rhog$.
\end{lemma}
\begin{proof}
  By induction on $\der \Gamma$.
\begin{caselist}

@@

>>>>

\begin{lemma}[Inversion of typing]
\label{lem:invty} \bla
\begin{enumerate}
\item If $\Gamma \der \lambda t : T'$ then either $\Gamma \der \piT U T \leq T'$ and
  $\cext \Gamma U \der t : T$ for some $U,T$ or $\Gamma \der \epiT\Size T \leq T'$ and
  $\eext \Gamma \Size \der t : T$ for some $T$.
\end{enumerate}
\end{lemma}

@@

We can now define the normalization relation
$\fbox{$\nbe \Gamma T t \evalsto v$} \iff \R {\Gamma} {\down{T \rho_\Gamma} (t \rho_\Gamma)} v$.
Whenever $\nbe \Gamma T t \evalsto v$, we may write $\nbe \Gamma T t$ for $v$.

@@

\begin{lemma}[Identity environment]
  \label{lem:idenv}
  If $\der \Gamma$ then
  $\LRS \Gamma \sid \Gamma \rhog$.
\end{lemma}
\begin{proof}
  By induction on $\der \Gamma$.
\begin{caselist}

@@

\begin{corollary}[Soundness of NbE]
  \label{cor:soundnbe} \bla
  \begin{enumerate}
  \item \label{it:sound}
  If\/ $\Gamma \der t : T$ then $\Gamma \der t = \nbe \Gamma T t : T$.
  \item
  If\/ $\Gamma \der t,t' : T$ and $\nbe \Gamma T t \equiv \nbe \Gamma T {t'}$ then
  $\Gamma \der t = t' : T$.
  \end{enumerate}
\end{corollary}
\begin{proof}
  \bla
  \begin{enumerate}

  \item For the identity environment $\LRS \Gamma \sid \Gamma \rhog$ (Lemm~\ref{lem:idenv})
  the Fundamental Theorem for Typing gives $\LR \Gamma t T {t\rhog} {T\rhog}$.
  This implies $\R \Gamma {\down{(T\rhog)} (t\rhog)} v$ for some normal form $v$ and
  $\Gamma \der t = v : T$ by Thm.~\ref{thm:into}.

  \item By transitivity of judgmental equality from (\ref{it:sound}).
     Here we use that the computed normal forms are identical because
     we erase all size annotations $\ann\alpha$ to $\ann\infty$ during readback.
  \end{enumerate}
\end{proof}

\begin{corollary}[Decidability of judgemental equality]
  If\/ $\Gamma \der t,t' : T$ then the test whether $\nbe\Gamma T t \equiv \nbe\Gamma T{t'}$
  terminates and
  decides $\Gamma \der t = t' : T$.
\end{corollary}
\begin{proof}
  Follows directly from soundness (which includes termination) and completeness of NbE.
\end{proof}

For normalization of types, we introduce \fbox{$\Nbe \Gamma T$}
via the relation $\Nbe \Gamma T \evalsto V \defas \RTy \Gamma {T\rhog} V$.
Analogously, we show that the test $\Nbe \Gamma T \equiv \Nbe \Gamma T'$
decides type equality $\Gamma \der T = T'$.

\begin{corollary}[Function type injectivity]
  \label{cor:piinj}
  \bla
  \begin{enumerate}
  \item \label{it:piinj}
    If $\Gamma \der \piT U T = \piT{U'}{T'} : s$ then
    $\Gamma \der U = U' : s$ and
    $\cext \Gamma U \der T = T' : s$.
  \item \label{it:allinj}
    If $\Gamma \der \epiT \Size T = \piT{\Size}{T'} : s$ then
    $\cext \Gamma \Size \der T = T' : s$.
  \end{enumerate}
\end{corollary}
\begin{proof}\bla
We prove statement (\ref{it:piinj}), the second statement follows analogously.
By the fundamental theorem,
$(\piT U T)\rhog = (\piT{U'}{T'})\rhog \in s$ which by definition means
$U\rhog = U'\rhog \in s$ and $T(\rhog,u) = T'(\rhog,u') \in s$ for all $u = u' \in s$.
By reification (Thm.~\ref{thm:rere}),
$\RTy\Gamma {U\rhog} V$ and $\RTy\Gamma {U'\rhg} V$ for some normal form
$V$.
Since by inversion we have $\Gamma \der U,U' : s$ we get $\Gamma \der U = U' : s$ by soundness of NbE.
Now, choosing $u := u' := \up U \var\Gamma \in U$, obtain
$T\rhogu = T' \rhogu \in s$, which analogously gives us $\cext\Gamma U \der T = T' : s$.
\end{proof}

The injectivity of function types allows a stronger inversion theorem for redexes.

\begin{lemma}[Strong inversion of redexes]
  \label{lem:stronginv}
  \bla
  \begin{enumerate}
  \item   If $\Gamma \der (\lambda t)\,u : T'$ then
  $\cext \Gamma U \der t : T$ and $\Gamma \der u : U$ and $\Gamma \der T[u] \leq T'$
  for some $U,T$.
  \item   If $\Gamma \der (\lambda t)\,a : T'$ then
  $\cext \Gamma \Size \der t : T$ and $\Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
  for some $U,T$.
  \item   If $\Gamma \der (\lambda t)\,\ann a : T'$ then
  $\erext \Gamma \Size \der t : T$ and $\resurrect \Gamma \der a : \Size$ and $\Gamma \der T[a] \leq T'$
  for some $U,T$.
  \end{enumerate}
\end{lemma}

\begin{theorem}[Subject reduction]
  \label{thm:sr}
  If $\Gamma \der t : T$ and $t \evalsto w$ then $\Gamma \der t = w : T$.
\end{theorem}
\begin{proof}
  By induction on $t \evalsto w$, using strong inversion on $\Gamma \der t : T$.
\end{proof}

\section{Type Checking}
\label{sec:tycheck}

Algorithmic subtyping \fbox{$\Gamma \der T \asub T'$}.
\begin{gather*}
  \ru{T \evalsto n \qquad
      T' \evalsto n' \qquad
      \Nbe \Gamma n \equiv \Nbe \Gamma n'
    }{\Gamma \der T \asub T'}
\\[2ex]
  \rux{T \evalsto \Set[\ell] \qquad
      T' \evalsto \Set[\ell']
    }{\Gamma \der T \asub T'
    }{\ell \leq \ell'}
\qquad
  \ru{T \evalsto \Nat a \qquad
      T' \evalsto \Nat a'
    }{\Gamma \der T \asub T'
    }{a \leq a'}
\\[2ex]
  \ru{T_1' \evalsto \piT{U_1}{T_1} \qquad
      T_2' \evalsto \piT{U_2}{T_2} \qquad
      \Gamma \der U_2 \asub U_1 \qquad
      \cext \Gamma {U_2} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\\[2ex]
  \ru{T_1' \evalsto \epiT\Size{T_1} \qquad
      T_2' \evalsto \epiT\Size{T_2} \qquad
      \cext \Gamma {\Size} \der T_1 \asub T_2
    }{\Gamma \der T_1' \asub T_2'}
\end{gather*}

\begin{lemma}[Soundness of algorithmic subtyping]
  If $\Gamma \der T \asub T'$ then $\Gamma \der T \leq T'$.
\end{lemma}
\begin{proof}
  By induction on $\Gamma \der T \asub T'$, using soundness of NbE and subject reduction.
\end{proof}
